---
title: "Time Series Modelling "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error = TRUE,message=FALSE, warning=FALSE)
options(width = 250)
suppressWarnings(suppressMessages(library(radiant)))
```

  
    
    
### Inferences from a Time Series Model, estimate a linear trend model and Include seasonal effects. Make predictions using the model developed.


```{r}
rm(list = ls())
cat("\f")
```


```{r}

rm(list = ls())
cat("\f")

# Install packages if you have never installed them.
# install.packages("xlsx")
# install.packages("rJava")
# install.packages("tseries")
# install.packages("forecast")

#Loading packages
library(tseries)
library(rJava)
library(xlsx)
library(lubridate)
```

```{r}

datas1 = read.xlsx('Keeling_CO2data_2017.xlsx',1,rowIndex = 8:713,colIndex = 1:5);
colnames(datas1) <- c("Yr","month","Date1","Date2","CO2.ppm")
# datas1 <- datas1[12:nrow(datas1),]

datas1$Yr <- ifelse(datas1$month == 12,datas1$Yr-1,datas1$Yr )

datas1$datenew <- mdy(paste(datas1$month,'01',datas1$Yr,sep = '-'))
datas1 <- datas1[,c(5:6)]
datas1$month <- as.factor(month(datas1$datenew))
datas1$month <- factor(datas1$month, levels=unique(datas1$month))

```


## CO2 PPM Rate  
## Part 1  

Plot the CO2 time series in column E. Briefly summarize the evidence of seasonal effects and trends in the time-series plot.  

```{r}
# Plot raw data
plot(datas1$datenew, datas1$CO2.ppm, xlab = 'date', ylab = 'co2', type='l')

mytimeseries = ts(data = datas1$CO2.ppm, start=c(1958, 2), end=c(2006, 10), frequency = 12)
plot(mytimeseries)
class(mytimeseries)

plot(datas1$CO2.ppm, type='l')
abline(reg=lm(datas1$CO2.ppm~time(datas1$CO2.ppm)))


# decompose(mytimeseries, "multiplicative") # decompose is in R base
plot(decompose(mytimeseries, type = "multiplicative"))
plot(aggregate(mytimeseries,FUN=mean))
#Box plot across months will give us a sense on seasonal effect
boxplot(mytimeseries~cycle(mytimeseries))
```


Solution 1 -  
Important Inferences  
1.	The year on year trend clearly shows that the amount of CO2 have been increasing without fail.  
2.	The mean value starts increasing in March, reaches its peak in April and then falls until August and then is more or less consistent for the next 4-5 months.  
3.	The variance of CO2 contents is consistent across all months.  
4.	The mean value of each month is a bit different and their variance is negligible. Hence, we have seasonal effect across months.  




## Part 2
Using data up to 2005, estimate a linear trend model for the CO2 measurements. Is the trend significant? Is the linear trend a good specification that yields reliable forecasts?    

```{r}
datas1.2005 <- datas1[datas1$datenew<= '2005-12-01',]
datas1.2005$month <- factor(datas1.2005$month, levels=unique(datas1.2005$month))
str(datas1.2005)
reg=lm(datas1.2005$CO2.ppm~time(datas1.2005$CO2.ppm))
summary(reg)
r <- reg$residuals # get the residuals from the fit
plot.ts(r)
Box.test(reg$residuals)
```

Solution 2-    

Yes the trend is significant . Looking at the statistical significance of the intercepts andcoefficient , we can say that there is a positive trend(Coefficient estimate for the time stamp is postive)of CO2 PPM level over time.However there is a clear trend in the residuals plot as it tends to have non constant variance and is heteroscedastic; so this is clearly not the best model and there is scope of improvement to make the model better, we can include inverse/squared terms in our regression and check if we get a model with no/little trend in the residuals.  

## Part 3
Using data up to 2005, develop a better trend specification. Report your trend estimates and explain what your specification is and what makes it better.  

```{r}
datas1.2005$CO2.ppm.sq <- datas1.2005$CO2.ppm^2
datas1.2005$CO2.ppm.in <- 1/datas1.2005$CO2.ppm^2
datas1.2005$CO2.ppm.cube <- datas1.2005$CO2.ppm^3
datas1.2005$time <- I(1:length(datas1.2005$CO2.ppm))
datas1.2005$time.sq <- I(datas1.2005$time^2)
datas1.2005$time.cube <- I(datas1.2005$time^3)

str(datas1.2005)
reg2=lm(CO2.ppm.in~time, data=datas1.2005)
summary(reg2)
r1 <- reg2$residuals
plot.ts(r1)
Box.test(reg2$residuals)
```
The residual plot indicates having no trend and all the coefficients estimates are statistically significant as the p values are less than .05 . The r squared value is improved when compared with the first model.    
Our Specification is to introduce an inverse of the dependent variable i.e CO2 PPM . We earlier saw the residual plot in the previous case to be inverted. So an inverted dependent variable will take care of that.  

## Part 4  
Include seasonal effects and report results for your preferred model that includes both trend and seasonal effects. Again, use data only up to 2005.  

```{r}
reg3=lm(CO2.ppm.in~time+ month, data=datas1.2005)
summary(reg3)
r1 <- reg3$residuals
plot.ts(r1)
Box.test(reg3$residuals)
```

Results from this model indicate an increase in Adjusted R-Square to 99.66% from 98.48%  


## Part 5
Using data up to 2005m12, forecast future CO2 for the period 2006m01 to 2016m10. Evaluate how good the forecasts from this model are. Are there any obvious problems with your forecasts [hint: are the forecast errors unpredictable]?  
```{r}
str(datas1)
datas1.2006 <- datas1[datas1$datenew > '2005-12-01',]
datas1.2006$month <- factor(datas1.2006$month, levels=unique(datas1.2006$month))
str(datas1.2006)
datas1.2006$CO2.ppm.sq <- datas1.2006$CO2.ppm^2
datas1.2006$CO2.ppm.in <- 1/datas1.2006$CO2.ppm^2
datas1.2006$CO2.ppm.cube <- datas1.2006$CO2.ppm^3
datas1.2006$time <- c(1:length(datas1.2006$CO2.ppm))
datas1.2006$time <- datas1.2006$time+575
datas1.2006$time.sq <- I(datas1.2006$time^2)
datas1.2006$time.cube <- I(datas1.2006$time^3)

# datas1.2006 <- datas1.2006[,c("time","month")]
# length(datas1.2005$CO2.ppm)


# df <- as.data.frame(time(datas1.2005$CO2.ppm),datas1.2006$month)
# pred.val <- predict.lm(reg3,newdata = as.data.frame(time(datas1.2005$CO2.ppm),datas1.2006[,c("month")])

pred.val <- predict(reg3,newdata = datas1.2006[,c("time","month")])

plot(datas1.2005$datenew, 1/(datas1.2005$CO2.ppm.in)^0.5, xlab = 'time', ylab = 'CO2 -PPM', type='l',col='green')
lines(datas1.2006$datenew, 1/(pred.val)^0.5,col='red')
plot(datas1.2006$datenew, 1/(pred.val)^0.5, col='red',type='l')

predicted.error <- 1/(datas1.2006$CO2.ppm.in)^0.5- 1/(pred.val)^0.5
plot.ts(predicted.error) 

```
  
the errors are almost constant .though it has in increasing trend towards the later part of the
predicted months  

```{r}
library(forecast)

accuracy(1/(datas1.2006$CO2.ppm.in)^0.5,1/(pred.val)^0.5)
plot(datas1.2006$datenew, 1/(datas1.2006$CO2.ppm.in)^0.5, xlab = 'time', ylab = 'CO2 -PPM', type='l',col='green')
lines(datas1.2006$datenew, 1/(pred.val)^0.5,col='red')

```

The predictions has been close for the most of the months and the RMSE for the predicted data is 2.4. The months that has low CO2 PPM has the highest deviation in prediction.  


## Part 6  
We believe that Population of Hawaai could be an important feature that could explain the variance in the data. Therefore we donwloaded the population data from this Source -

https://united-states.reaproject.org/analysis/comparative-trends-analysis/population/reports/150000/0/

```{r}
# browseURL("https://united-states.reaproject.org/analysis/comparative-trends-analysis/population/reports/150000/0/")
```


```{r}
library(tidyverse)
Hawaii_population = read.xlsx('Hawaai_Population.xlsx',1);
datas1.2006$Year <- year(datas1.2006$datenew)
datas1.2006.new <- datas1.2006 %>% left_join(Hawaii_population,by="Year") 

datas1.2005$Year <- year(datas1.2005$datenew)
datas1.2005.new <- datas1.2005 %>% left_join(Hawaii_population,by="Year") 

reg4=lm(CO2.ppm.in~time+ month + Population, data=datas1.2005.new)
summary(reg4)
r1 <- reg4$residuals
plot.ts(r1)
Box.test(reg4$residuals)

pred.val.new <- predict(reg4,newdata = datas1.2006.new[,c("CO2.ppm.in","time","month","Population")])

plot(datas1.2005.new$datenew, 1/(datas1.2005.new$CO2.ppm.in)^0.5, xlab = 'time', ylab = 'CO2 -PPM', type='l',col='green')
lines(datas1.2006.new$datenew, 1/(pred.val.new)^0.5,col='red')
plot(datas1.2006.new$datenew, 1/(pred.val.new)^0.5, col='red',type='l')

predicted.error <- 1/(datas1.2006.new$CO2.ppm.in)^0.5- 1/(pred.val)^0.5
plot.ts(predicted.error) 
```
  
the errors are almost constant .though it has in increasing trend towards the later part of the
predicted months  

```{r}
library(forecast)
accuracy(1/(datas1.2006$CO2.ppm.in)^0.5,1/(pred.val)^0.5)
accuracy(1/(datas1.2006.new$CO2.ppm.in)^0.5,1/(pred.val.new)^0.5)
plot(datas1.2006.new$datenew, 1/(datas1.2006.new$CO2.ppm.in)^0.5, xlab = 'time', ylab = 'CO2 -PPM', type='l',col='green')
lines(datas1.2006.new$datenew, 1/(pred.val.new)^0.5,col='red')
```

As we can see that introducing the Population of Hawaai as one of the component, we are able
to generate better prediction accuracy even on the Test/unseen data. The RMSE has gone down from
2.4 to 2.11, a drop of ~16%.  

## Part 7  

```{r}
No.of.mon.frm.2005 <- 12*(2020-2005) +575
December.2020 <- data.frame(time = No.of.mon.frm.2005, month = '12')

December.2020.CO2.ppm <- 1/(predict(reg3,newdata = December.2020))^0.5
December.2020.CO2.ppm

```
The prediction for December 2020 is 410.76. This forecast might not be accurate due to the increasing error term in the model that needs to be addressed.  



